Flat histogram method comparison on 2D Ising Model

### 2019-11-23 papers/histogram/sad-referee-response-2.md

Dear editor,

We appreciate the referees' careful reading of our paper, and feel that their
feedback has led to a significant improvement. We have added new data and a
section for the 31 atom Lennard-Jones cluster. We show specific heat curves for
this test system and show the convergence SAD vs the other method quantitatively
(a quantitative comparison of convergence on the LJ31 system has not been done in the literature to
our knowledge).  We also added a note to the Square-well fluid section
highlighting that this system has not been explored using broad histogram
methods. We believe that we have addressed all of the referees' concerns and per
the first referee's own words 'If that or similar data would be in the paper
there would be no doubt whatsoever that SAD is a "substantial improvement of
existing computational methods".'

@@

Response to first referee
# -----------------------------------------------------------------------------#

In the previous review, the first referee acknowledged that  'The paper is very
well written. It is also true that the authors addressed all of my previous
concerns in a satisfactory manner.'

The first referee asked that we simulate (either 'faster and/or reliably')
low-temperature solid-solid transitions in specific Lennard-Jones clusters. The
referee gave LJ135 as an example, but stated that if specific heat capacity
curves were in the paper for LJ135 or 'similar' (we did LJ31 for reasons we
state in the paper) then 'there would be no doubt whatsoever that SAD is a
"substantial improvement of existing computational methods".'
We believe that we have completely satisfied the first referee and thank the referee
for pointing our attention toward simulating Lennard-Jones clusters.

We were surprised by how poorly the 1/t-WL algorithm performed on the LJ31 system,
in particular the convergence was very strongly affected by the lower limit of
the energy range chosen.  This example thus strongly highlights the benefits
SAD provides when examining the heat capacity over a chosen temperature range.

### 2020-01-24 papers/histogram/sad-referee-response-2.md

Dear editor,

We appreciate the referees' careful reading of our paper, and feel
that their feedback has led to a significant improvement. We have
added new data and a section for the 31 atom Lennard-Jones cluster. We
compute the low-temperature specific heat for this test system and
show the convergence SAD vs the other methods quantitatively. A
quantitative comparison of convergence on the LJ31 system has not been
done in the literature to our knowledge.  We also added a note to the
Square-well fluid section highlighting that this system has not
previously been explored using broad histogram methods. We believe
that we have addressed all of the referees' concerns and per the first
referee's own words 'If that or similar data would be in the paper
there would be no doubt whatsoever that SAD is a "substantial
improvement of existing computational methods".'

@@

...

Response to first referee
# -----------------------------------------------------------------------------#

In the previous review, the first referee acknowledged that  'The paper is very
well written. It is also true that the authors addressed all of my previous
concerns in a satisfactory manner.'

The first referee asked that we simulate (either 'faster and/or
reliably') low-temperature solid-solid transitions in specific
Lennard-Jones clusters. The referee gave LJ135 as an example, but
stated that if specific heat capacity curves were in the paper for
LJ135 or 'similar' then 'there would be no doubt whatsoever that SAD
is a "substantial improvement of existing computational methods".' We
chose to examine the LJ31 cluster for reasons we state in the paper,
specifically that this cluster has a clear solid-solid phase
transition at very low temperature that has been shown in the past to
be challenging to simulate using flat histogram methods.  We believe
that we have completely addressed the request by the first referee and
thank the referee for pointing our attention toward simulating
Lennard-Jones clusters.

We were surprised by how poorly the 1/t-WL algorithm performed on the
LJ31 system.  In particular, the convergence was very strongly
affected by the lower limit of the energy range chosen.  This example
thus strongly highlights the benefits SAD provides when examining the
low-temperature heat capacity, where the temperature of interest is
known, but the range of energies that affect that temperature are hard
to determine.

Response to second referee
# -----------------------------------------------------------------------------#
All of the second referee's comments have been addressed previously.

### 2020-01-24 papers/histogram/ising.tex

Liang independently considered whether WL could be treated as a special case of
Stochastic Approximation whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang et
al.~\cite{liang2007stochastic} argued that WL can be considered a form of
Stochastic Approximation Monte Carlo (SAMC). Unlike WL, SAMC can guarantee
convergence (if certain conditions are met). Despite the added benefit of
guaranteed convergence, the method still has a system specific user-defined
variable. Such variables often create difficulty when applying Monte-Carlo
methods across arbitrary systems.  The SAD (Stochastic Approximation with
Dynamic $\gamma$) method outlined by Pommerenck
et.al~\cite{pommerenck2019stochastic} is a special version of the SAMC algorithm
that dynamically chooses the modification factor rather than relying on system
dependent parameters. SAD shares the same convergence properties with SAMC while
replacing un-physical user-defined parameters with the algorithms dynamic
choice.  In this work, we apply the family of weight-based flat histogram Monte
Carlo methods (WL, 1/t-WL, SAMC, SAD) to the 2D Ising model.

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact} make the
system sufficiently interesting for such theoretical comparisons. It is
also of important note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the infinite volume limit ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over all enumerated spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
where $g(E)$ is the multiplicity of the system which is identical to
$D(E)$. We can estimate the average deviation from the exact solution
using the relation~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control}:
\begin{align}
\braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
\end{align}
As per the implementation by Wang and
Landau~\cite{wang2001determining}, the total number of available energy
states are $N-1$.
% \jpsays{Also the necessary condition that the ground state is
% $2$ for the 2D Ising model is enforced to guarantee the accuracy of the
% density of states at low temperatures. If $2^N$ states is the condition
% used for the ferromagnetic ising model, the rescaled factor is
% determined primarily by the maximum density of states impacing the
% accuracy of the energy levels at or near the ground state.}
%\section{Discrete Spin Systems} In this work, we compare the
%convergence properties of the family of Weight-Based Broad Histogram
%Methods applied to the Q-State Potts model and Ising model. The two
%classical spin systems are ideal testbeds for comparing the convergence
%of Monte-Carlo methods. The Potts and Ising models have 1st and 2nd
%order phase transitions respectively and will both require energy
%binning due to the discrete energy levels of the systems. Convergence
%scaling for the family of Weight-Based Broad Histogram methods can
%easily be explored on the two non-continuous systems.

@@

>>>>

Liang independently considered whether WL could be treated as a special case of
Stochastic Approximation whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang et
al.~\cite{liang2007stochastic} argued that WL can be considered a form of
Stochastic Approximation Monte Carlo (SAMC). Unlike WL, SAMC can guarantee
convergence (if certain conditions are met). Despite the added benefit of
guaranteed convergence, the method still has a system specific user-defined
variable. Such variables often create difficulty when applying Monte-Carlo
methods across arbitrary systems.  The SAD (Stochastic Approximation with
Dynamic $\gamma$) method outlined by Pommerenck
et.al~\cite{pommerenck2019stochastic} is a special version of the SAMC algorithm
that dynamically chooses the modification factor rather than relying on system
dependent parameters. SAD shares the same convergence properties with SAMC while
replacing un-physical user-defined parameters with the algorithms dynamic
choice.

In this work, we apply the family of weight-based flat histogram Monte
Carlo methods (WL, 1/t-WL, SAMC, SAD) to the 2D Ising model.

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can estimate the average deviation from the exact solution
using the relation~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control}:
\begin{align}
\braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
\end{align}
As per the implementation by Wang and
Landau~\cite{wang2001determining}, the total number of available energy
states are $N-1$.
% \jpsays{Also the necessary condition that the ground state is
% $2$ for the 2D Ising model is enforced to guarantee the accuracy of the
% density of states at low temperatures. If $2^N$ states is the condition
% used for the ferromagnetic ising model, the rescaled factor is
% determined primarily by the maximum density of states impacing the
% accuracy of the energy levels at or near the ground state.}
%\section{Discrete Spin Systems} In this work, we compare the
%convergence properties of the family of Weight-Based Broad Histogram
%Methods applied to the Q-State Potts model and Ising model. The two
%classical spin systems are ideal testbeds for comparing the convergence
%of Monte-Carlo methods. The Potts and Ising models have 1st and 2nd
%order phase transitions respectively and will both require energy
%binning due to the discrete energy levels of the systems. Convergence
%scaling for the family of Weight-Based Broad Histogram methods can
%easily be explored on the two non-continuous systems.

@@

\davidsays{The results feel a bit slim.  I wonder whether we can look
  at the heat capacity also? Or perhaps there is a way that we can see
  something ``interesting'' arise out of the computation? I also
  wonder if we can run longer and get a bit more accuracy.}
